# Tests and inferences {#intro}

Definitely the first thing to be familiar with while doing machine learning works is the basic of statistical inferences.  
In this chapter, we go over some of the few important topics and r-ways to do them.  

Let's get started.  

## T-tests
The __independent t test__ is used to test if there is any statistically *significant difference between two means*. Use of an independent t test requires several assumptions to be satisfied.  

1.  The variables are continuous and independent
1.  The variables are normally distributed
1.  The variances in each group are equal

When these assumptions are satisfied the results of the t test are valid. Otherwise they are invalid and you need to use a non-parametric test. When data is not normally distributed you can apply transformations to make it normally distributed.

Using the `mtcars` data set, we check if there are any difference in mile per gallon (mpg) for each of the automatic and manual group.  

Check the data and mark as factor the driving system. 

```{r intro01, message=FALSE}
library(tidyverse)
glimpse(mtcars)
df <- mtcars
df$am <- factor(df$am, labels = c("automatic", "manual"))
df2 <- df %>% select(mpg, am)
glimpse(df2)
```

Generate descriptive statistic for each group.  

```{r intr02}
df2 %>% group_by(am) %>% 
  summarise(mean = mean(mpg), minimum = min(mpg), maximum = max(mpg))
```

Generate boxplot for each group
```{r intro03_boxplot}
ggplot(df2, aes(x = am, y = mpg)) + 
  geom_boxplot(fill = c("dodgerblue3", "goldenrod2")) + 
  labs(title = "Achieved milage for Automatic / Manual cars", 
       x = "Type of car")
```

Test the normality of the data.  
To do so, we can use the __Shapiro Wilk Normality Test__
```{r intro04_summary}
df2 %>% group_by(am) %>% 
  summarise(shaprio_test = shapiro.test(mpg)$p.value)

```

There is no evidence of departure from normality.  

Test the equal variance in each group.  
To do so, we use the `levene.test` from the `car` package.  
```{r intro05_leveneTest}
car::leveneTest(mpg ~ am, center = mean, data = df2)
```

Because the variance in the 2 groups is not equal, we have to transform the data.  

Apply a log transformation to stabilize the variance.  
```{r intro06_logtransformed}
log_transformed_mpg = log(df2$mpg)
```

Now we can finally apply the t test to our data.  
```{r intro07_ttest}
t.test(log_transformed_mpg ~ df2$am, var.equal = TRUE)
```

Interpret the results.  

* Manual cars have on average a higher mileage per Gallon (24 mpg)  compared to Automatic (17 mpg).  
* The box plot did not reveal the presence of outliers
* The Shapiro-Wilk normality test did not show any deviation from normality in the data  
* The Levene Test showed difference in the variance in the 2 group.  We addressed that by log transforming the data  
* The t test show a significant difference in the mean of miles per gallon from automatic and manual cars.  
